DRAFT: Harmonized and trait aggregation paper

### 2020-10-20 Paper/Draft_harmonization_paper.tex
Updates from Overleaf

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Species traits, defined as measurable properties of an organism \cite{mcgill_rebuilding_2006}, might be beneficial in achieving this goal \cite{heino_jani_macroecological_2013}. %?just traits 
Traits evolve through adaptations (e.g., physiological, behavioural, etc.) of organisms to their environment and indicate direct or indirect linkages between the biological response of an organism to its environment \cite{southwood_habitat_1977, verberk_delivering_2013}.
% TODO: PUP: Not only „of an organism“, but also „of a population“ (at least for some of the traits related to reproduction; e.g. difficult to describe „voltinism“ as the organism scale!).
Besides providing a mechanistic explanation of species-environment relationships, trait-based approaches might be suitable for large scale analysis because the variability in trait responses is lower than in taxonomic responses \cite{bonada_taxonomic_2007, baird_toward_2011}. Traits of freshwater invertebrate individuals are difficult to determine because - unlike plants - their traits often cannot be measured directly.
For example, to gain knowledge on feeding habits requires evaluating mouthpart morphology, consumed food, and the organisms function within its community \cite{moog_comprehensive_nodate}. 
% PUP says: But the trait-based approach can benefit from two centuries of detailed, published observations on biological traits and ecological preferences made by zoologists/ecologists on a large range of species in various environmental conditions.
Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors or as trait metrics in biomonitoring \cite{poff_developing_2010, szocs_effects_2014, bhowmik_large_2015, menezes_beyond_2010}.

In the last decades, freshwater ecologists developed comprehensive invertebrate trait databases for various biogeographic regions \cite{usseglio-polatera_biomonitoring_2000, schmidt-kloiber_www.freshwaterecology.info_2015, vieira_database_nodate, Philips_and_Smith_NZ_DB_2018, kefford_integrated_2020, tomanova_trophic_2006}. The availability of invertebrate trait data from different biogeographic regions enables comparisons of trait variation and their relation to environmental factors across large scales. However, such analyses have been carried out mostly within biogeographic regions, using information from one or two trait databases. For example, Bonada et al. \cite{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe using traits from Usseglio-Polatera et al. \cite{usseglio-polatera_biomonitoring_2000}, Poff et al. \cite{poff_developing_2010} characterized trait composition across sites in the Western US using traits from Poff et al. \cite{poff_functional_2006}, and Botwe et al. \cite{botwe_effects_2018} investigated the effect of salinity on invertebrate traits in different sites in South Australia using trait data from Poff et al. \cite{poff_functional_2006} and Schäfer et al. \cite{schafer_trait_2011}. 
% ? Include Saito for South America 
Analyses of invertebrate traits that synthesize information on invertebrate grouping features from more than two different biogeographic regions are rare. A grouping feature is defined by Schmera et al. as a general property (e.g. feeding mode) that comprises a "group of related traits (e.g., predator, shredder, etc.) that vary among species or among individuals within a species" \cite{schmera_proposed_2015}. To our knowledge, only Brown et al. \cite{brown_functional_2018} harmonized grouping features from more than two geographically distant invertebrate trait databases in a study on the influence of decreasing glacier cover on functional diversity and community assembly of invertebrates. 
% TODO: ?Paper trait convergence Doledec

We suspect that the heterogeneity of information in freshwater invertebrate trait databases is likely a major reason for the lack of studies across biogeographic regions. To harmonize grouping features from different regions, first commonly accepted and unambiguous trait definitions are required \cite{schneider_towards_2019}. In the best case, grouping features would be classified into the same traits across databases or they could easily be harmonized using standardized terminology. However, a lack of standardized terminology of trait definitions and poor metadata quality in many trait databases is a common issue throughout the field of trait-based ecology \cite{baird_toward_2011, schneider_towards_2019}. Secondly, consistent coding of traits facilitates the compatibility of trait data from different databases. Traits can be binary (i.e. trait exists or not), continuous, or fuzzy coded variables. Fuzzy codes represent the affinity of an organism to express a certain trait. They are used to account for plasticity in traits and are usually converted to percentages. Continuous values are typically used for grouping features that can be measured, like body size. However, invertebrate trait databases are heterogeneous with regard to the coding they use for their traits \cite{culp_incorporating_2011} (Table \ref{tab:trait_databases_coding_differentiation}). Brown et al. \cite{brown_functional_2018} harmonized grouping features based on trait databases from Europe, North America, and New Zealand because in these trait databases identical grouping features are classified differently into traits. As the traits from North America were coded binary in contrast to the fuzzy coded traits from Europe and New Zealand, the authors consulted experts to assign fuzzy coded traits to North American taxa or inferred them from the European trait database. Thus, it becomes apparent that using invertebrate trait data from several regions requires extensive data processing prior to the actual data analysis. A centralized database with standardized and unambiguous traits and a consistent coding of traits would minimize data processing effort.

@@

Aggregated trait affinities of the five trait aggregation methods (\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg}) were compared to trait affinities assigned at family-level by experts, which were available for the Australian and North American database for a subset of grouping features and taxa. For the Australian database, we compared aggregated trait affinities with assigned trait affinities resolved at family-level for the grouping features feeding mode and size using data from Chessman \cite{chessman_dissolved-oxygen_2018}. In Chessman \cite{chessman_dissolved-oxygen_2018} feeding mode is classified similarly as in the harmonized Australian database except that the trait parasite is missing. We conducted the comparison for the 220 families available in Chessman \cite{chessman_dissolved-oxygen_2018}. Considering each factor combination of family and trait (in total 8) this amounts to 1760 cases.

@@

%! Presenting results of trait analysis using harmonized traits:
\subfile{Results/Results_trait_harmonization}
% TODO: Fix reference to SI!

@@

% Influence of aggregation methods on trait-environment relationships:
% 1) In general, differences produced by the aggregation methods seem to arise by using mean or median, 
% not driven by different weighting approaches (direct vs. stepwise vs. weighted) 
% -> link to taxonomic hierarchy, many families have only a few genera
% -> Similarly, many genera have only a few species
% -> linear relationship: the more genus per family, the more species (although there are exceptions -> e.g. Europe)
% 
% Differences for NOA greater due to coding?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Ben's Comment: -> But another issues is that these aggregation methods were being compared to traits assigned at the family level, which I understand was often done by expert opinion. So, these traits assigned at the family level may not be really “correct”! The difference between the trait aggregated to family and its designation at the family level could indicate: (a) the aggregation method is “no good”, (b) the trait assigned at the family level was “no good”, or (c) both a and b.

% Sampled from a truncated normal distribution values for three traits
% (upper and lower boundaries 0 and 1)
% Traits were normalized so that they sum up to one
% Three different taxonomic hierarchies were simulated: 
% Simulated one family with 25 species 
% -> 1) 5 Genera:  5 Species respectively
% -> 2) 5 Genera: 1 Genus 13 Species (more than ~ 50 %); 4 Genera with 3 species
% -> 3) 5 Genera: All have different amount of species

% Simulation result: 
% -  base simulation with similar amount of speices per genus and genera per family and varying trait values
% results can vary when variation in trait values increases 
% However, mostly the different aggregation methods result in similar trait affinities
% when differences occur, only for: 
% dir_median vs stepwise median
% direct mean vs stepwise median
% stepwise mean vs stepwise median
% direct mean vs direct median
% stepwise median vs weighed agg
% Trait variability: 0.3 and greater
% Max diff ~ 0.2

% - Results for varying taxonomic hierarchy:
% Which levels?
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% 2) Harmonization: Minor influence on RDA species scores
% RDA species scores 
%   -> Normalized eigenvectors
%   -> dependent on nr. of variables? Nr. of response variables not really influential 
% We used only a small-scale study -> What's different to large-scale studies?

% Further aggr. after harmonization: no difference to the harmonized dataset
% ?Logical because trait aggr. methods themselves showed only minor differences

% Complex Stat. techniques developed -> cannot use full potential

@@

>>>>

Explaining and predicting how communities are shaped by environmental factors is one of the main goals of ecology. Organismal traits, defined as measurable properties of an organism \cite{mcgill_rebuilding_2006}, might be beneficial in achieving this goal \cite{heino_jani_macroecological_2013}. Traits evolve through adaptations (e.g., physiological, behavioral, etc.) of organisms to their environment and indicate direct or indirect linkages between the biological response of an organism to its environment \cite{southwood_habitat_1977, verberk_delivering_2013}. Besides providing a mechanistic explanation of species-environment relationships, trait-based approaches might be suitable for large scale analysis because the variability in trait responses is lower than for taxonomic responses \cite{bonada_taxonomic_2007, baird_toward_2011}. Traits of freshwater invertebrates are difficult to determine because, unlike plants, their traits often cannot be measured directly. For example, to gain knowledge on feeding habits requires evaluating mouthpart morphology, consumed food, and the organisms function within its community \cite{moog_comprehensive_nodate}. Nevertheless, invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors or as trait metrics in biomonitoring \cite{poff_developing_2010, szocs_effects_2014, bhowmik_large_2015, menezes_beyond_2010}.

%%% Old Intro Version %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Explaining and predicting how aquatic communities are shaped by environmental factors is one of the main goals of freshwater ecology. Organismal traits, defined as measurable properties of an organism \cite{mcgill_rebuilding_2006}, are increasingly incorporated into freshwater ecology to support this goal, e.g. by relating macroinvertebrate trait composition to environmental factors or as trait metrics in biomonitoring \cite{poff_developing_2010, szocs_effects_2014, bhowmik_large_2015, menezes_beyond_2010}. Traits evolve through adaptations (e.g., physiological, behavioral, etc.) of organisms to their environment and indicate direct or indirect linkages between the biological response of an organism to its environment \cite{southwood_habitat_1977, verberk_delivering_2013}. 
% Besides providing a mechanistic explanation of species-environment relationships, trait-based approaches may be suitable for large scale analysis because the variability in trait responses is lower than for taxonomic responses \cite{bonada_taxonomic_2007, baird_toward_2011}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the last decades, freshwater ecologists developed comprehensive invertebrate trait databases for various biogeographic regions \cite{usseglio-polatera_biomonitoring_2000, schmidt-kloiber_www.freshwaterecology.info_2015, vieira_database_nodate, Philips_and_Smith_NZ_DB_2018, kefford_integrated_2020, tomanova_trophic_2006}. The availability of invertebrate trait data from different biogeographic regions enables comparisons of trait variation and their relation to environmental factors across large scales. However, such analyses have been carried out mostly within biogeographic regions, using information from one or two trait databases. For example, Bonada et al. \cite{bonada_taxonomic_2007} compared trait composition for mediterranean and temperate regions in Europe using traits from Usseglio-Polatera et al. \cite{usseglio-polatera_biomonitoring_2000}, Poff et al. \cite{poff_developing_2010} characterized trait composition across sites in the Western US using traits from Poff et al. \cite{poff_functional_2006}, and Botwe et al. \cite{botwe_effects_2018} investigated the effect of salinity on invertebrate traits in different sites in South Australia using trait data from Poff et al. \cite{poff_functional_2006} and Schäfer et al. \cite{schafer_trait_2011}. 
% ? Include Saito for South America 
Analyses of invertebrate traits that synthesize information on invertebrate grouping features from more than two different biogeographic regions are rare. A grouping feature is defined by Schmera et al. as a general property (e.g. feeding mode) that comprises a "group of related traits (e.g., predator, shredder, etc.) that vary among species or among individuals within a species" \cite{schmera_proposed_2015}. To our knowledge, only Brown et al. \cite{brown_functional_2018} harmonized grouping features from more than two geographically distant invertebrate trait databases in a study on the influence of decreasing glacier cover on functional diversity and community assembly of invertebrates. 
% TODO: ?Paper trait convergence Doledec

We suspect that the heterogeneity of information in freshwater invertebrate trait databases is likely a major reason for the lack of studies across biogeographic regions. To harmonize grouping features from different regions, first commonly accepted and unambiguous trait definitions are required \cite{schneider_towards_2019}. In the best case, grouping features would be classified into the same traits across databases or they could easily be harmonized using standardized terminology. However, a lack of standardized terminology of trait definitions and poor metadata quality in many trait databases is a common issue throughout the field of trait-based ecology \cite{baird_toward_2011, schneider_towards_2019}. Secondly, consistent coding of traits facilitates the compatibility of trait data from different databases. Traits can be binary (i.e. trait exists or not), continuous, or fuzzy coded variables. Fuzzy codes represent the affinity of an organism to express a certain trait. They are used to account for plasticity in traits and are usually converted to percentages. Continuous values are typically used for grouping features that can be measured, like body size. However, invertebrate trait databases are heterogeneous with regard to the coding they use for their traits \cite{culp_incorporating_2011} (Table \ref{tab:trait_databases_coding_differentiation}). Brown et al. \cite{brown_functional_2018} harmonized grouping features based on trait databases from Europe, North America, and New Zealand because in these trait databases identical grouping features are classified differently into traits. As the traits from North America were coded binary in contrast to the fuzzy coded traits from from Europe and New Zealand, the authors consulted experts to assign fuzzy coded traits to North American taxa or inferred them from the European trait database. Thus, it becomes apparent that using invertebrate trait data from several regions requires extensive data processing prior to the actual data analysis. A centralized database with standardized and unambiguous traits and a consistent coding of traits would minimize data processing effort.

@@

Aggregated trait affinities of the five trait aggregation methods (\textit{direct\_agg\textsubscript{median}}, \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{median}}, \textit{stepwise\_agg\textsubscript{mean}}, and \textit{weighted\_agg}) were compared to trait affinities assigned at family-level by experts, which were available for the Australian and North American database for a subset of grouping features and taxa. For the Australian database, we compared aggregated trait affinities with assigned trait affinities resolved at family-level for the grouping features feeding mode and size using data from Chessman et al. \cite{chessman_dissolved-oxygen_2018}. In Chessman et al. \cite{chessman_dissolved-oxygen_2018} feeding mode is classified similarly as in the harmonized Australian database except that the trait parasite is missing. We conducted the comparison for the 220 families available in Chessman et al. \cite{chessman_dissolved-oxygen_2018}. Considering each factor combination of family and trait (in total 8) this amounts to 1760 cases.

@@

%! Presenting results of trait analysis using harmonized traits:
\subfile{Results/Results_trait_harmonization}

@@

% In general, differences produced by the aggregation methods seem to arise by using mean or median, not driven by different weighting approaches (direct vs. stepwise vs. weighted) 
% Complex Stat. techniques developed -> cannot use full potential

@@

% Trait definitions 
\subfile{Trait_def_table/SI_trait_def_table2.tex}
